### 2020-01-17 manuscript/manuscript.rmd
update the manuscript

Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of orthologs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig. 2). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices. Moreover, the little overlap among studies may potentially lead to incongruent phylogenetic results and difficult to directly compare results across studies.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

```{r support_published_analyses}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```

**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).

@@

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60+F. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig. 5). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

@@

![](figures/Figure_Pairwise_heatmap_animal_root.png)
**Extended Data Fig. 3.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance. (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses. (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses. The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation. The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].

@@

## The current state of understanding and future direction
Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices. By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases. Three factors contribute significantly to the previous incongruent results. First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies. Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models. More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgoups. Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species. Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister. Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.

@@

Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, a consensus set of orthologs and analytical strateges should be generated and formalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

#### Phylogenetic analyses in IQtree
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C models in the model testing. For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.  For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloByes analyses were run for at least one month.

#### phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious]. For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.

>>>>

Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated since all these studies rely on their own method for orthology inference. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of orthologs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig. 2). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices. Moreover, the little overlap among studies may potentially lead to incongruent phylogenetic results and difficult to directly compare results across studies.

@@

```{r support_published_analyses}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```
**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

@@

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig. 5). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

@@

![](figures/Figure_Pairwise_heatmap_animal_root.png)
**Extended Data Fig. 3.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance. (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses. (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses. The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation. The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].

@@

## The current state of understanding and future direction
Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices. By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases. Three factors contribute significantly to the previous incongruent results based on our analyses. First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies. Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models. More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice. Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species. Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister. Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.

@@

Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

#### Phylogenetic analyses in IQtree
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C models in the model testing. For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.  For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloByes analyses were run for at least one month.

#### phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious]. For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
